Title

### 2022-04-05 thesis/experiments/experiments.tex
add journals and restructured

\section{Frequency control} \label {sec: controle_freq}

@@

After an idle time, the processor begins to activate C states, starting with C0 where it is still fully active, then moving to C1, where some features are disabled, up to Cn, where all possible features are disabled. The ACPI standard establishes the functionalities that can be disabled between level C1 and level C3, as seen in Table \ref {tab: states_c}. The other levels are specific to each manufacturer.

\begin {table} [H]
\centering
\begin {tabular} {| l | l | l |}
\hline
Mode & Name & Functionality \\ \hline
C0 & operating state & Active processor \\ \hline
C1 & Halt & Stop executing instructions \\ \hline
C2 & Stop-Clock & Disable the internal clock \\ \hline
C3 & Sleep & Disable cache coherence \\ \hline
\end {tabular}
\caption {C states}
\label {tab: states_c}
\end {table}

In state C, the higher the level the greater the energy savings, but returning to the fully functional level is more difficult. In states P, there is a trade-off between performance and energy savings. Figure \ref {fig: p_state} best illustrates the change of states, in which we can see which parts of the circuits are deactivated in states C, the latency to return to the active state, power consumption and also shows the relationship of states P with often.

\begin {figure} [H]
\centering
\begin{subfigure}[t]{0.5\textwidth}
\centering
\includegraphics[height= 7.5cm]{intro/Imagens/c_states}
\caption {C states}
\end {subfigure}%
~
\begin{subfigure}[t]{0.5\textwidth}
\centering
\includegraphics[height= 7.5cm]{intro/Imagens/p_states}
\caption {States P}
\end {subfigure}
\caption {Illustration of states C and P} {Altered image from \protect \url {https://www.thomas-krenn.com/en/wiki/Processor_P-states_and_C-states}}
\label {fig: p_state}
\end {figure}

@@

On Linux there are several frequency management options \cite {Brown2005}. The main ones are acpi-cpufreq, Intel P-state, AMD powernow. In this work, acpi-cpufreq is used, which is standard and allows direct control of frequency through system files. Acpi-cpufreq is a Linux module that uses implemented policies that dynamically decide the frequency to be used. Some of these policies are:

\begin {itemize}
\item Performance - configured as often as possible
\item Powersave - configured as low as possible
\item Userspace - the user chooses the frequency to be used
\item Ondemand - controls the frequency depending on the processor load. When the load increases the frequency also increases accordingly.
\item Conservative - similar to Ondemand but more smoothly, the frequency increase is continuous instead of jumping.
\end {itemize}

\section {Power consumption monitoring} \label {sec: monitoring}
The Running Average Power Limit (RAPL) and Intelligent Platform Management Interface (IPMI) interfaces were used to measure the power consumed. described in \cite{November2013}.

\subsection {IPMI}
IPMI \cite {November2013} is a set of specifications for autonomous subsystems that provides processor, firmware and operating system independent management and monitoring. The use of IPMI allows system administrators to avoid having to travel to the server location, which is often far away, to perform their tasks. Also, servers are located in places with low temperature and with a lot of noise due to the ventilation system, and one should avoid spending too much time in these places. With remote management, it is possible to turn the system on and off, remotely access the (BIOS) and reinstall the system in case of any serious failure.

\begin {figure} [H]
\centering
\includegraphics [height = 7.5cm] {intro/Imagens/IPMI-Block-Diagram.png}
\caption {IMPI diagram} {Image taken from \protect \url {https://pt.wikipedia.org/wiki/Intelligent_Platform_Management_Interface}, the main components of IPMI and how they communicate are shown}
\label {fig: IPMI}
\end {figure}

@@

\subsection {RAPL}

Modern Intel microprocessors, based on the SandyBridge architecture, include the RAPL \cite {Rotem2012, Hahnel2012, Hackenberg2015} interface designed to limit the use of energy on a chip while ensuring maximum performance. This interface supports energy measurement capabilities through an integrated circuit that estimates energy use based on a model driven by architectural event counters for all components. It also provides temperature readings and current leak models. Estimates are made available in model-specific registers (MSR), updated in milliseconds. The energy estimates offered by RAPL were validated by Intel, which showed excellent results.

\section{Verifying models hypothesis}

\subsection{Frequency voltage ratio}

\begin{table}[H]
	\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|c|}
		\hline
		Freq (GHz) & 2.2   & 2.1   & 2.0   & 1.9   & 1.8   & 1.7   & 1.6   & 1.5   & 1.4   & 1.3   & 1.2   \\ \hline
		V          & 0.77  & 0.76  & 0.75  & 0.74  & 0.73  & 0.72  & 0.71  & 0.70  & 0.69  & 0.68  & 0.67  \\ \hline
		aperf      & 2.199 & 2.099 & 2.000 & 1.899 & 1.799 & 1.699 & 1.599 & 1.500 & 1.397 & 1.297 & 1.200 \\ \hline
	\end{tabular}
\end{table}

\begin{figure}[H]
	\centering
	\includegraphics[width=\columnwidth]{experiments/figures/freq_volt_rel.png}
	\caption{Frequency voltage relation}
	\label{fig:freq_volt_rel}
\end{figure}

\subsection{Instructions executed}

\begin{table}[H]
	\caption{Cores variation}
	\begin{tabular}{|c|c|c|c|}
		\hline
		Application  & Instructions & Dev.     & Dev. (\%) \\ \hline
		Vip          & 7.97e+11     & 7.16e+06 & 0.00\%    \\ \hline
		Openmc       & 8.17e+07     & 1.65e+04 & 0.02\%    \\ \hline
		Rtview       & 9.91e+12     & 1.55e+09 & 0.02\%    \\ \hline
		X264         & 4.52e+11     & 5.81e+07 & 0.01\%    \\ \hline
		Bodytrack    & 1.86e+12     & 3.95e+10 & 2.13\%    \\ \hline
		Fluidanimate & 2.09e+12     & 8.44e+10 & 4.04\%    \\ \hline
		Xhpl         & 1.14e+08     & 1.24e+05 & 0.11\%    \\ \hline
		Blackschole  & 3.75e+12     & 1.40e+09 & 0.04\%    \\ \hline
		Dedup        & 1.02e+11     & 5.74e+07 & 0.06\%    \\ \hline
		Swapti       & 2.43e+12     & 8.87e+08 & 0.04\%    \\ \hline
		Canneal      & 1.19e+11     & 4.46e+07 & 0.04\%    \\ \hline
		Freqmine     & 1.27e+12     & 4.78e+08 & 0.04\%    \\ \hline
		Ferret       & 4.76e+11     & 7.04e+07 & 0.01\%    \\ \hline
	\end{tabular}
\end{table}

\begin{table}[H]
	\caption{Frequency Variation}
	\begin{tabular}{|c|c|c|c|}
		\hline
		Application  & Instructions & Dev.     & Dev. (\%) \\ \hline
		Vip          & 7.97e+11     & 1.16e+06 & 0.00\%    \\ \hline
		Openmc       & 8.17e+07     & 4.52e+03 & 0.01\%    \\ \hline
		Rtview       & 9.91e+12     & 6.64e+05 & 0.00\%    \\ \hline
		X264         & 4.52e+11     & 1.54e+05 & 0.00\%    \\ \hline
		Bodytrack    & 1.84e+12     & 2.54e+05 & 0.00\%    \\ \hline
		Fluidanimate & 2.38e+12     & 1.70e+09 & 0.07\%    \\ \hline
		Xhpl         & 1.14e+08     & 5.95e+03 & 0.01\%    \\ \hline
		Blackschole  & 3.75e+12     & 4.36e+05 & 0.00\%    \\ \hline
		Dedup        & 1.02e+11     & 8.32e+07 & 0.08\%    \\ \hline
		Swapti       & 2.43e+12     & 1.48e+05 & 0.00\%    \\ \hline
		Canneal      & 1.19e+11     & 3.01e+05 & 0.00\%    \\ \hline
		Freqmine     & 1.27e+12     & 3.70e+08 & 0.03\%    \\ \hline
		Ferret       & 4.76e+11     & 5.63e+07 & 0.01\%    \\ \hline
	\end{tabular}
\end{table}

\subsection{Input size and instructions}

\begin{figure}[H]
	\centering
	\includegraphics[width=\columnwidth]{models/figures/hypothesis/input_instructions/fp/Blackschole.png}
	\caption{Caption}
	\label{fig:my_label}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=\columnwidth]{models/figures/hypothesis/input_instructions/fp/Canneal.png}
	\caption{Caption}
	\label{fig:my_label}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=\columnwidth]{models/figures/hypothesis/input_instructions/input_time/Blackschole.png}
	\caption{Caption}
	\label{fig:my_label}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=\columnwidth]{models/figures/hypothesis/input_instructions/input_time/Canneal.png}
	\caption{Caption}
	\label{fig:my_label}
\end{figure}

\section{Model experimental validation} \label{sec:experimentalvalidation}
In this section, the models presented in~\cref{sec:powermodel} and~\cref{sec:performancemodel} were validated with a benchmark specific for multi-core architectures. Additionally, in order to assess the modeling overhead and accuracy, our proposal was then compared to machine learning approaches. We compared against Support Vector Regression (SVR)~\cite{Smola2004}, Decision Tree \cite{Kitts2006RegressionLecture}, k-nearest neighbors \cite{Altman1992AnRegression}, Multilayer perceptron \cite{Murtagh1991MultilayerRegression}, and some new methods such as Gao et al. \cite{Gao2019DendriticPrediction}. However, SVR was chosen as the most representative because in our tests it performed best without an aggressive fine-tuning.

\subsection{Case-Study Applications} \label{sec:casestudyapplication}
The PARSEC parallel benchmark suite, version 3.0~\cite{Bienia2008}, OpenMC \cite{Romano2015OpenMC:Development} and LINPACK (HPL) \cite{Dongarra1988TheExplanation}, were chosen as case studies. The PARSEC benchmark focused on emerging workloads and was designed to represent the next-generation shared-memory programs for chip-multiprocessors. It covers an ample range of areas such as financial analysis, computer vision, engineering, enterprise storage, animation, similarity search, data mining, machine learning, and media processing. The OpenMC and the LINPACK are two classic HPC programs.

\subsection{Case-Study Architecture} \label{sec:casestudyarchitecture}
The experiments were executed in one computer node equipped with two Intel Xeon E5-2698 v3 processors with sixteen cores each and two hardware threads for each core. The overall view of the architecture is shown in Figure \ref{fig:architecture}. The maximum non-turbo frequency is 2.3GHz, and the total physical memory of the node is 128GB (8$\times$16GB). Turbo frequency and hardware multi-threading were disabled during all experiments. The operating system used was Linux CentOS 6.5, kernel 4.16.

\begin{figure}[htb!]
	\centering
	\includegraphics[width=\columnwidth]{models/figures/architecture.png}
	\caption{Node architecture (the image was made with the lstop application).}
	\label{fig:architecture}
\end{figure}

The Linux kernel has many different policies for power management, depending on the driver. In the default driver, the acpi-cpufreq, the options are:
\begin{itemize}
	\item Powersave
	\item Performance
	\item Ondemand
	\item Conservative
	\item Userspace
\end{itemize}
In this work, the frequency control was performed using the Userspace governor, and the core control was accomplished by modifying the appropriate system files with the default CPU-hotplug driver.

The architecture is equipped with the Intelligent Platform Management Interface (IPMI), a set of interfaces allowing out-of-band management of computer systems and platform-status monitoring via the local network~\cite{Schwenkler2006IntelligentInterface}. It can monitor variables and resources such as the system's temperature, voltage, fans, and power supplies, with independent sensors attached to the hardware.

\subsection{Fitting the Models} \label{sec:fitting}
To find the parameters of the \cref{eq:en_final}, 10 uniformly random configurations of frequencies ($f$), cores ($p$) and inputs ($N$) were chosen from the range $1<=p<=32$, $1.2<=f<=2.2$ and $1<=N<=5$ respectively. The application was executed for each chosen configuration, and the measured  energy and time values were collected. For the input size if we assume that all CPU instructions are executed at approximately the same time, the number of basic operations will be directly correlated with the time. Thus, we can estimate the problem size by looking at the execution time, allowing us to divide a large problem size into several smaller ones knowing their relationship as done in the work of Oliveira ~\cite{Oliveira2018ApplicationCores}. The unity can also vary depending on the definition. In our case for simplicity, we assign numbers from 1 the smallest size to 10 the largest, increasing the problem linearly, that way its also possible to interpolate any input in between these values.

%The application's input was chosen to increase the workload linearly, from smallest to largest, so that most of the input space is covered~\cite{Oliveira2018ApplicationCores}.

For each configuration, samples of the power were collected using IPMI every 1 second. This sampling rate was chosen because the order of magnitude of the mean run time of the applications is minutes. Therefore, this rate provides enough samples to measure average power. Additionally, timestamps and the total run time were collected. The total energy spent on each configuration is estimated by first interpolation the power samples using a first order method and then integrating this function in the time.

From the sampled data, the parameters of the model can be estimated. This results in an optimization problem of finding the parameters that minimize the distance between estimated and the measured values. To solve this minimization problem, non-linear least squares have been applied.

The python library scikit-learning was used to build the SVR model~\cite{PedregosaFandVaroquauxGandGramfortAandMichel2011Scikit-learn:Python}. The SVR was trained using the same data used for parameters estimation of \cref{eq:en_final} with a grid search used to find the best kernel function and the best values for the hyper-parameters penalty for the wrong ($C$) and ($\gamma$). For this data, the best function was the Radial Base Function (RBF), and the hyper-parameters were $C=10^4$ and $\gamma=0.5$.

\subsection{Measured versus Modeled Energy}
\label{sec:measuredversusmodeledenergy}

To validate the proposed energy model, all possible configurations were tested by varying the cores in a range of $1<=p<=32$, the frequency in $1.2<=f<=2.2$ and the input in $1<=N<=5$, which varies from 400 to over 1000 configurations depending on the application as some applications have restrictions on the number of cores. The mean percentage error was computed as the difference between the estimated and measured values according to the following equation:
\begin{equation}
MPE = \frac{1}{N} \sum_i^N \frac{|y_{\rm estimated}-y_{\rm measured}|}{y_{\rm measured}}.
\label{eq:mpe}
\end{equation}

\subsubsection{Frequency x Cores}
Figures \ref{fig:en_eq_black}, \ref{fig:en_eq_canneal} plot the measured and modeled energy consumption for some of the applications modeled. They  show some of the possible shapes that the model can take while varying the number of active cores, operating frequency, and input size.
\begin{figure}[H]
	\centering
	\begin{subfigure}[b]{0.48\textwidth}
		\centerline{\includegraphics[width=\columnwidth]{models/figures/energy/freq_cores/completo_black_5.png}}
		\caption{Blackscholes.}
		\label{fig:en_eq_black}
	\end{subfigure}
	%
	\begin{subfigure}[b]{0.48\textwidth}
		\centerline{\includegraphics[width=\columnwidth]{models/figures/energy/freq_cores/completo_canneal_1.png}}
		\caption{Canneal.}
		\label{fig:en_eq_canneal}
	\end{subfigure}
	\caption{Example fit for a specific input size: Blackscholes (a) and Canneal (b).  “measured values” are the sensor data and “minimum energy” is the minimum energy model prediction.
	}
\end{figure}
\subsubsection{Frequency x Input}
Figures \ref{fig:en_eq_black}, \ref{fig:en_eq_canneal} plot the measured and modeled energy consumption for some of the applications modeled. They  show some of the possible shapes that the model can take while varying the number of active cores, operating frequency, and input size.
\begin{figure}[H]
	\centering
	\begin{subfigure}[b]{0.48\textwidth}
		\centerline{\includegraphics[width=\columnwidth]{models/figures/energy/freq_inps/completo_black_5.png}}
		\caption{Blackscholes.}
		\label{fig:en_eq_black}
	\end{subfigure}
	%
	\begin{subfigure}[b]{0.48\textwidth}
		\centerline{\includegraphics[width=\columnwidth]{models/figures/energy/freq_inps/completo_canneal_1.png}}
		\caption{Canneal.}
		\label{fig:en_eq_canneal}
	\end{subfigure}
	\caption{Example fit for a specific input size: Blackscholes (a) and Canneal (b).  “measured values” are the sensor data and “minimum energy” is the minimum energy model prediction.
	}
\end{figure}
\subsubsection{Cores x Input}
Figures \ref{fig:en_eq_black}, \ref{fig:en_eq_canneal} plot the measured and modeled energy consumption for some of the applications modeled. They  show some of the possible shapes that the model can take while varying the number of active cores, operating frequency, and input size.
\begin{figure}[H]
	\centering
	\begin{subfigure}[b]{0.48\textwidth}
		\centerline{\includegraphics[width=\columnwidth]{models/figures/energy/cores_inps/completo_black_5.png}}
		\caption{Blackscholes.}
		\label{fig:en_eq_black}
	\end{subfigure}
	%
	\begin{subfigure}[b]{0.48\textwidth}
		\centerline{\includegraphics[width=\columnwidth]{models/figures/energy/cores_inps/completo_canneal_1.png}}
		\caption{Canneal.}
		\label{fig:en_eq_canneal}
	\end{subfigure}
	\caption{Example fit for a specific input size: Blackscholes (a) and Canneal (b).  “measured values” are the sensor data and “minimum energy” is the minimum energy model prediction.
	}
\end{figure}

>>>>

\section{Frequency control} \label{sec:frequency_control}

@@

After an idle time, the processor begins to activate C states, starting with C0 where it is still fully active, then moving to C1, where some features are disabled, up to Cn, where all possible features are disabled. The ACPI standard establishes the functionalities that can be disabled between level C1 and level C3, as seen in Table \ref {tab:states_c}. The other levels are specific to each manufacturer.

\begin{table}[H]
\centering
\begin{tabular}{| l | l | l |}
\hline
Mode & Name & Functionality \\ \hline
C0 & operating state & Active processor \\ \hline
C1 & Halt & Stop executing instructions \\ \hline
C2 & Stop-Clock & Disable the internal clock \\ \hline
C3 & Sleep & Disable cache coherence \\ \hline
\end{tabular}
\caption{C states}
\label{tab:states_c}
\end{table}

In state C, the higher the level the greater the energy savings, but returning to the fully functional level is more difficult. In states P, there is a trade-off between performance and energy savings. \cref{fig:p_state} best illustrates the change of states, in which we can see which parts of the circuits are deactivated in states C, the latency to return to the active state, power consumption and also shows the relationship of states P with often.

\begin {figure} [H]
\centering
\begin{subfigure}[t]{0.5\textwidth}
\centering
\includegraphics[width=\columnwidth]{intro/figures/c_states}
\caption{C states}
\end{subfigure}%
~
\begin{subfigure}[t]{0.5\textwidth}
\centering
\includegraphics[width=\columnwidth]{intro/figures/p_states}
\caption{States P}
\end{subfigure}
\caption {Illustration of states C and P} {Altered image from \protect \url {https://www.thomas-krenn.com/en/wiki/Processor_P-states_and_C-states}}
\label{fig:p_state}
\end{figure}

@@

On Linux there are several frequency management options \cite {Brown2005ACPILinux}. The main ones are acpi-cpufreq, Intel P-state, AMD powernow. In this work, acpi-cpufreq is used, which is standard and allows direct control of frequency through system files. Acpi-cpufreq is a Linux module that uses implemented policies that dynamically decide the frequency to be used. Some of these policies are:

\begin{itemize}
\item Performance - configured as often as possible
\item Powersave - configured as low as possible
\item Userspace - the user chooses the frequency to be used
\item Ondemand - controls the frequency depending on the processor load. When the load increases the frequency also increases accordingly.
\item Conservative - similar to Ondemand but more smoothly, the frequency increase is continuous instead of jumping.
\end{itemize}

\section{Power consumption monitoring} \label{sec:power_consumption_monitoring}
The Running Average Power Limit (RAPL) and Intelligent Platform Management Interface (IPMI) interfaces were used to measure the power consumed. described in \cite{IPMI2013ConfigurationGuide}.

\subsection{IPMI}
IPMI \cite{IPMI2013ConfigurationGuide} is a set of specifications for autonomous subsystems that provides processor, firmware and operating system independent management and monitoring. The use of IPMI allows system administrators to avoid having to travel to the server location, which is often far away, to perform their tasks. Also, servers are located in places with low temperature and with a lot of noise due to the ventilation system, and one should avoid spending too much time in these places. With remote management, it is possible to turn the system on and off, remotely access the (BIOS) and reinstall the system in case of any serious failure.

\begin{figure}[H]
\centering
\includegraphics[height = 7.5cm] {intro/figures/IPMI-Block-Diagram.png}
\caption{IMPI diagram} {Image taken from \protect \url {https://pt.wikipedia.org/wiki/Intelligent_Platform_Management_Interface}, the main components of IPMI and how they communicate are shown}
\label{fig:IPMI}
\end{figure}

@@

\subsection{RAPL}

Modern Intel microprocessors, based on the SandyBridge architecture, include the RAPL \cite {Rotem2012Power-managementBridge, Hahnel2012RAPL, Hackenberg2015AnProcessor} interface designed to limit the use of energy on a chip while ensuring maximum performance. This interface supports energy measurement capabilities through an integrated circuit that estimates energy use based on a model driven by architectural event counters for all components. It also provides temperature readings and current leak models. Estimates are made available in model-specific registers (MSR), updated in milliseconds. The energy estimates offered by RAPL were validated by Intel, which showed excellent results.

\section{Case-Study Architecture} \label{sec:casestudyarchitecture}
The experiments were executed in one computer node equipped with two Intel Xeon E5-2698 v3 processors with sixteen cores each and two hardware threads for each core. 
The overall view of the architecture is shown in \cref{fig:architecture}.
The maximum non-turbo frequency was 2.3 GHz, and the total physical memory of the node was 128 GB (8 $\times$ 16 GB). Turbo frequency and hardware multi-threading were disabled during all experiments. The operating system used was Linux CentOS 6.5, kernel 4.16.

\begin{figure}[H]
	\centering
	\includegraphics[width=\columnwidth]{models/figures/architecture.png}
	\caption{Node architecture (the image was made with the lstop application).}
	\label{fig:architecture}
\end{figure}

The Linux kernel has many different policies for power management, depending on the driver. In the default driver, the acpi-cpufreq, the options are Powersave, Performance, Ondemand, Conservative, and Userspace. Each governor has a policy on how the frequency is selected. In this investigation, the frequency control was performed using the Userspace governor, which allows the user or any userspace program to set the CPU to a specific frequency. The core control was accomplished by modifying the appropriate system files with the default CPU-hotplug driver.

The architecture was equipped with the intelligent platform management interface (IPMI), a set of interfaces allowing out-of-band management of computer systems and platform-status monitoring via the local network~\cite{Schwenkler2006IntelligentInterface}. It can monitor variables and resources, such as the system's temperature, voltage, fans, and power supplies, with independent sensors attached to the hardware.

\section{Case-Study Applications} \label{sec:casestudyapplication}
The applications blackscholes, bodytrack, canneal, dedup, fluidanimate, freqmine, raytrace, swaptions, vips and x264 from the PARSEC \url{https://parsec.cs.princeton.edu/download.htm} (accessed on 20 February  2020) parallel benchmark suite, version 3.0~\cite{Bienia2008TheSuite}, OpenMC \cite{Romano2015OpenMC:Development} and LINPACK (HPL) \cite{Dongarra1988TheExplanation}, were chosen as case studies. The PARSEC benchmark focused on emerging workloads and was designed to represent the next-generation shared-memory programs for chip-multiprocessors. It covers an ample range of areas, such as financial analysis, computer vision, engineering, enterprise storage, animation, similarity search, data mining, machine learning, and media processing. The OpenMC and the LINPACK are two classic HPC programs.

%The other two experiments use applications Raytrace~\cite{Bienia2008} and OpenMC~\cite{Romano2015OpenMC:Development}. Raytrace is an application addressed on rendering photorealistic scenes. It is available on the PARSEC Benchmark suite, version 2.0. The PARSEC benchmark, focused on emerging workloads, represents the next-generation shared-memory programs for chip multiprocessors. It covers a wide range of areas such as financial analysis, computer vision, engineering, enterprise storage, animation, similarity search, data mining, machine learning, and media~processing.
%
%\textls[-20]{{OpenMC is an application that implements the Monte-Carlo method to simulate the transport of neutrons and photons. It is a classic program aimed at high-performance computing.}}

\section{Validation the framework}\label{sec:validation_the_framework}
In this Section, we discuss the measurements and simulation results for three distinct cases, including two real applications. We also demonstrate how external libraries and standalone visualization tools can be used to render the collected data and investigate performance aspects such as the program's scalability capacity.

\subsection{Configuration} \label{subsec:configuration_pascal}

We used three experiments to assess the tool and demonstrate its ability to support analysis aimed at observing parallel scalability. The first includes a runtime imbalance between processing units in a specific parallel region. In this case, the objective is to present how the analyzer helps users observe the impact of the inefficiency of a code part on the whole program. The code for this experiment is presented in \cref{lst:codes_regionscomparison}. It consists of two simple parallel regions with the same functionality that divide the iterations of a loop between the available threads. The difference between the regions lies in the strategies used to manipulate the sum variable, used in the example to store the values of the calculations performed in each thread. We assume that the {\tt anyop()} function, in this case, invariably has the same runtime in all function calls.

The command presented on \cref{lst:pascal_usage_ex} was used as base to perform experiments. For the experiment of \cref{lst:codes_regionscomparison}, we use just the parameters {\tt -c}, {\tt -r}, {\tt -t}, {\tt -a} and {\tt -o}, including the 64 value to {\tt -c} option.

\lstset{style=ccodestyle, frame=tb}
\begin{lstlisting}[label={lst:pascal_usage_ex}, language=bash, caption={Command line showing how experiments were run through a terminal.}]
	analyzer 
	application
	-c 1,2,4,8,...,32 # threads/cores
	-r 3 # number of repetitions
	-t auto # automated instrumentation
	-a 1 # aggregation mode
	--ipmi ip user psswd # energy sensor
	--idtm 5 # idle time between runs
	--dhpt # disable hyper-thread
	--dcrs # disable cores
	--ipts ... # application specific inputs
	-o application.json # output file
\end{lstlisting}

\lstset{style=ccodestyle, frame=tb}
\begin{lstlisting}[label={lst:codes_regionscomparison}, language=C, caption={Sample code used to visualize the impact of regions on program scalability.}]
	int main(int argc, char **argv) {
		unsigned long sum = 0;
		int ops = atoi(argv[1]);

		#pragma omp parallel for schedule(static) reduction(+: sum)
		for (int i = 0; i < ops; i++)
		sum += anyop();

		#pragma omp parallel for schedule(static)
		for (int i = 0; i < ops; i++) {
			#pragma omp critical
			sum += anyop();
		}
	}
\end{lstlisting}

The command return a .json file with all the information necessary for our analysis. We can quickly generate tables with the collected data from this file, thus supporting observation and analysis of scalability, energy trends, and model fitting. The code described in \cref{lst:regtab} and \cref{tab:regtable_dedup} are samples of how users can easily read and view collected data from the control terminal in a tabular way.

\lstset{style=pythonStyle, frame=tb}
\begin{lstlisting}[label={lst:regtab}, language=python, captionpos=b, caption={Example of using the Python API to load analyzer files.}]
	from analyzer import Data

	data = Data("application.json")
	data.energy(regions=True)
	# data.speedup()
	# data.efficiency()
	# data.regions()
\end{lstlisting}

%The result can be seen in tabular form, as shown in Table.~\ref{tab:regtable_dedup}. Using these tables, we can quickly generate figures showing the application's behavior from various points of view. In Figures \ref{fig:speedup_raytrace} and \ref{fig:speedup_openmc} we observe the trend of the speedup with the number of cores and input.
\begin{table}[H]
	\caption{Dataframe generated automatically from collected samples using the Python API.}
	\label{tab:regtable_dedup}

	\resizebox{\columnwidth}{!}{%
		\begin{tabular}{ccccccc}
			\toprule
			\textbf{Repetition} & \textbf{Input} & \textbf{Cores} & \textbf{Regions} & \textbf{Start Time (s)} & \textbf{Stop Time (s)} & \multicolumn{1}{c}{\textbf{ipmi Energy (J)}} \\ \midrule
			1 & 1 & 1 & 1 & 0.00 & 66.31 & 13,156.19 \\
			1 & 1 & 1 & 2 & 0.00 & 66.27 & 13,148.87 \\
			... & ... & ... & ... & ... & ... & ... \\
			1 & 5 & 30 & 4 & 0.00 & 8.46 & 1903.82 \\
			1 & 5 & 30 & 5 & 0.02 & 58.16 & 13,067.01 \\ \bottomrule
	\end{tabular}}
\end{table}

\subsection{Pascal Analyzer Validation}\label{sec:results_pascal}

The analyzer does not display graphics and other visual elements natively. However, the simple use of external libraries allows generating graphics and visualizing points of the program's behavior that you want to observe. In addition, specialized visualization tools can also be used to view the results. PaScal Viewer \cite{Silva2018}, for example, natively interprets the analyzer's output files, complementing its functionality and creating an integrated and appropriate environment for the program's parallel scalability analysis.

To analyze the experiment presented in \cref{lst:codes_regionscomparison}, we use PaScal Viewer to observe in a hierarchical way how the different OpenMP clauses impacted the region's efficiencies. The efficiency pinpoints how a program can take advantage of increasing processing elements on a parallel system. It is defined as the ratio of speedup to the number of processing units. In this sense, PaScal Viewer displays an efficiency diagram for each analysis region, as presented in \cref{fig:pv_regionscomparison}. Comparing these diagrams, it is possible to see how the critical clause damages the scalability. In addition, it is also possible to relate that this second region affects the efficiency of the whole program due to the code characteristics.

If the reduction clause replaces {\tt \#pragma~omp~crtical}, the second region and the entire program become more efficient, as shown in \cref{fig:pv_regionscomparison_2}. The PaScal Viewer diagrams mentioned in this work, the x-axis (i1--i7) corresponds to different data inputs and the y-axis (2--64) to the number of threads used in the processing. Figures~\ref{fig:pv_regionscomparison} and \ref{fig:pv_regionscomparison_2} were rendered using a tool feature that smoothes the color transition of diagrams. This feature uses interpolation to create a visual element where the color transition is less pronounced. The diagram axes only show the initial and final values with this option. The user can visualize diagrams with only discrete values or even compare the two presentation modes side by side. \cref{fig:visualizationmodes} shows the difference between discrete and smoothed modes considering the whole program and the simulation scenario that uses the clause {\tt \#pragma~omp~crtical}.
\begin{figure}[H]
	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth]{pascalanalyzer/figures/results/regionscomparison_r1.pdf}
		\caption{\centering}
		\label{fig:pv_regionscomparison_a}
	\end{subfigure}
	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth]{pascalanalyzer/figures/results/regionscomparison_r2.pdf}
		\caption{\centering}
		\label{fig:pv_regionscomparison_b}
	\end{subfigure}
	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth]{pascalanalyzer/figures/results/regionscomparison_whole.pdf}
		\caption{\centering}
		\label{fig:pv_regionscomparison_c}
	\end{subfigure}
	\caption{Efficiency diagrams and impact of inner regions on program scalability. (x-axis = inputs; y-axis = number of threads). (\textbf{a}) First region diagram. (\textbf{b}) Second region diagram. (\textbf{c}) Whole program~diagram.}
	\label{fig:pv_regionscomparison}
\end{figure}
\unskip
\begin{figure}[H]
	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth]{pascalanalyzer/figures/results/efficiency_rg_1.pdf}
		\caption{\centering}
		\label{fig:pv_regionscomparison_a_2}
	\end{subfigure}
	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth]{pascalanalyzer/figures/results/efficiency_rg_2.pdf}
		\caption{\centering}
		\label{fig:pv_regionscomparison_b_2}
	\end{subfigure}
	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth]{pascalanalyzer/figures/results/efficiency_rg_3.pdf}
		\caption{\centering}
		\label{fig:pv_regionscomparison_c_2}
	\end{subfigure}
	\caption{Efficiency diagrams after removing {\tt \#pragma~omp~crtical} clause. (x-axis = inputs; y-axis = number of threads). (\textbf{a}) First region diagram. (\textbf{b}) Second region diagram. (\textbf{c}) Whole program~diagram.}
	\label{fig:pv_regionscomparison_2}
\end{figure}
\unskip
\begin{figure}[H]
	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth]{pascalanalyzer/figures/results/diagram_discretevalues.pdf}
		\caption{\centering}
		\label{fig:discretevalues}
	\end{subfigure}
	%
	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth]{pascalanalyzer/figures/results/diagram_smoothedvalues.pdf}
		\caption{\centering}
		\label{fig:smoothedvalues}
	\end{subfigure}

	\caption{Visualization modes of diagrams provided by PaScal Viewer. (\textbf{a}) PaScal Viewer diagram on discrete mode. (\textbf{b}) PaScal Viewer diagram on smoothed mode.}
	\label{fig:visualizationmodes}
\end{figure}

%The analyzer allows that the result can be seen in tabular form, as shown in Table.~\ref{tab:regtable_dedup}. Using these tables, we can quickly generate figures that show the application's behavior from various points of view. In Figures \ref{fig:speedup_raytrace} and \ref{fig:speedup_openmc} we can observe the trend of the speedup with the number of cores and input.

%Figures \ref{fig:raytrace_en} and \ref{fig:openmc_en} present the energy variation with the number of cores.
